TIMES Ireland Model: Residential Sector

### 2021-05-26 manuscript.tex
26 May2021

\begin{abstract}
%% Text of abstract

Write after paper is completed %Ireland's residential sector has particularly underachieved in previous climate targets.  Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector. The disadvantageous starting position has impeded residential decarbonisation progress thus far. \par
%Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy. \par
%The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.

@@

%% main text
\section{Introduction}
\label{sec:Introduction}
\subsection{Climate Targets}

@@

Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}. Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths. An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}. Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty. These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}. Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}. Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling. The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.
\subsection{Residential Energy Services}
\label{Intro:EnergyServices}

@@

The energy-related GHG emissions are expected to be reduced beyond the 51\% national target by 2030 compared to 2018, due to the higher cost-effectiveness of abating GHG emissions when compared with agriculture.  The CAP2019 Marginal Abatement Cost Curve (MACC), which was developed by McKinsey and Company's modelling tools and is based on EPA's GHG emission projection report \cite{EPA2018EPAReport} found electricity \& transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. Therefore a cost effective solution to decarbonising the residential sector, is to receive low carbon electricity and increasing the share of electricity usage.

@@

\subsubsection{Hydrotreated Vegetable Oil}
While it maybe optimal to use Hydrotreated Vegetable Oil (HVO) in transport, it is also considered in the residential sector, as it can provide a cheap alternative for the 700,000 Irish dwellings using home heating oil ( mostly kerosene). HVO which meets the EU sustainability criteria, reduce GHG emissions by 60\% as compared to fossil fuels \cite{Soam2019FactorsSweden}. The capacity of HVO is growing in Europe, but HVO is currently about twice the price of kerosene. HVO can be used to replace kerosene, albeit with cost of up to €400 per oil boiler, however there is no policy for HVO in the residential sector.

\section{Literature Review}
\label{LitR}
%PICKUPHERE
``How should Ireland set GHG targets in the residential sector?” 
Choosing the Pathways which Create the Least Burden
and Offer the Most Opportunity for Ireland
\subsection{Modelling Methodology or Approach}
\label{LitR:Model}
The TIM approach.....
\subsection{Supply}
\label{LitR:Supply}
\subsection{Demand}
\label{LitR:Demand}
\subsection{Assumptions}
\label{LitR:Assumptions}
 % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 
- LI,er.alThe nationwide survey identified existing technologies, age,
income, region, dwelling characteristics, and knowledge of eco- technology as the six most influential factors for determining homeowners' preferences for heating systems. Among
%the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings. The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances. There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs. Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030. The effect of these targets also stretches out across the construction sector. The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.  This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.

>>>>

\begin{abstract}
%% Text of abstract
%Ireland's residential sector has particularly underachieved in previous climate targets. Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector. The disadvantageous starting position has impeded residential decarbonisation progress thus far. \par
%Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy. \par
%The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
Write after paper is completed

@@

%% main text
\section{Introduction}
\label{sec:Introduction}

Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways, and interpreting the results. This section looks at Ireland's climate targets, energy policy and distinctive residential sector, before examining how researchers have used Energy system optimization models (ESOMs) to analyse the residential sector.
Section \ref{LitR} describes the methodology used to develop the residential sector of TIMES Ireland Model (TIM). The sector outlines the data used and calculations used ......
What does new model do that previous didn’t?
Does new policy context require new modelling approach?
How does new modelling structure provide insights (give examples) in a way that previous model didn’t?
% ``How should Ireland set GHG targets in the residential sector?” 
This is done by examining the current climate targets, EU dirictives and Irish and EU policy, then analysiing future policy and looking at ESOMs

\subsection{Climate Targets}

@@

Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}. Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths. An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}. Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty. These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}. Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}. Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling. The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.

@@

The CAP2019 Marginal Abatement Cost Curve (MACC),  \cite{EPA2018EPAReport} found electricity \& transport are the most cost effective for decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. Therefore a cost effective solution to decarbonising the residential sector, is to receive low carbon electricity and increasing the share of electricity usage.

@@

\subsubsection{Hydrotreated Vegetable Oil}
While it maybe optimal to use Hydrotreated Vegetable Oil (HVO) in transport, it is also considered in the residential sector, as it can provide a cheap alternative for the 700,000 Irish dwellings using home heating oil ( mostly kerosene). HVO which meets the EU sustainability criteria, reduce GHG emissions by 60\% as compared to fossil fuels \cite{Soam2019FactorsSweden}. The capacity of HVO is growing in Europe, but HVO is currently about twice the price of kerosene. HVO can be used to replace kerosene, albeit with cost of up to €400 per oil boiler, however there is no policy for HVO in the residential sector.
biofuel obligation scheme - Internationally there is a limited supply of HVO available and the
demand for the fuel is high.HVO has been produced in the past at Ireland’s only refinery at Whitegate and
potentially could be again with the appropriate investment. Not HVO100, but higher concentration of HVO than current blending capacity

\section{Literature Review}
\label{LitR}

``How should Ireland set GHG targets in the residential sector?” 
Choosing the Pathways which Create the Least Burden
and Offer the Most Opportunity for Ireland
\subsection{Modelling Methodology or Approach}
\label{LitR:Model}
The TIM approach.....
%a newly formed group with members from different universities, industries and semi-state agencies within Ireland.  Following this, other datasets were then used to develop the full housing stock, based on the correlation of year of construction and building energy ratings.
\subsection{Residential Energy Services}
\label{Intro:EnergyServices}

\subsection{Supply}
\label{LitR:Supply}
\subsection{Demand}
\label{LitR:Demand}
%Space Heating demand - BER space heating is only within certain months and assumes 21C heating..
\subsection{Assumptions}
\label{LitR:Assumptions}
Macroeconomic Assumptions 
Capacity
Costs
Efficiency
Savings
 % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 
- LI,er.alThe nationwide survey identified existing technologies, age,
income, region, dwelling characteristics, and knowledge of eco- technology as the six most influential factors for determining homeowners' preferences for heating systems. Among
%the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings. The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances. There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs. Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030. The effect of these targets also stretches out across the construction sector. The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.  This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.
